New short-term orbits of irregular satellites in order to predict stellar occultations up to 2020

### 2016-06-28 Irregular-occ/Irregular-occ.tex
.

@@

\title[Orbits and predictions for occultations up to 2020]{New orbits of irregular satellites for the predictions of stellar occultations up to 2020}
%\title[Orbits and predictions for occultations up to 2020]{New orbits of irregular satellites based on thousands of new observations and predictions for stellar occultations up to 2020}
\author[A. R. Gomes-J\'unior, M. Assafin, L. Beauvalet et al.]{A. R. Gomes-J\'unior$^{1}$\thanks{E-mail: altair08@astro.ufrj.br},
M. Assafin$^{1,\dag}$,
L. Beauvalet$^{2,3}$,
J. Desmars$^{4}$,\newauthor
R. Vieira-Martins$^{1,2,5,\dag}$,
J. I. B. Camargo$^{2,5}$,
B. E. Morgado$^{1,2}$
F. Braga-Ribas$^{2,6}$
\\
$^{1}$Observat\'orio do Valongo/UFRJ, Ladeira Pedro Ant\^onio 43,
CEP 20.080-090 Rio de Janeiro - RJ, Brazil\\
$^{2}$Observat\'orio Nacional/MCTI, R. General Jos\'e Cristino 77, CEP 20921-400 Rio de Janeiro - RJ, Brazil\\
$^{3}$Observatoire de Paris/SYRTE, 77 Avenue Denfert Rochereau 75014 Paris, France\\
$^{4}$Institut de m\'ecanique c\'eleste et de calcul des \'eph\'em\'erides - Observatoire de Paris, UMR 8028 du CNRS, 77 Av. Denfert-Rochereau,\\ 75014 Paris, France\\
$^{5}$Laborat\'orio Interinstitucional de e-Astronomia - LIneA, Rua Gal. Jos\'e Cristino 77, Rio de Janeiro, RJ 20921-400, Brazil\\
$^{6}$Federal University of Technology - Paran\'a (UTFPR / DAFIS), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil
}
\begin{document}
\newcommand{\noccs}{5442 } % total number of occultations
\defcitealias{GomesJunior2015}{GJ15}

@@

We use the following notations: \begin{itemize}
\item \textbf{in one dynamical family consisting of $\mathcal{N}$ irregular satellites, $i$ will stand for the one whose equation of motion we are considering, $l$ will stand for another irregular satellite in gravitationnal interaction belonging to the same family}
\item $J$ Jupiter
\item $j$ another body of the Solar System, \textbf{among the 10 included (Sun+planets+Pluto+Moon)}
\item $M_j$ the mass of the $j$ body, not an irregular satellite
\item $m_i$ the mass of the irregular satellite $i$
\item $\vec{r_i}$ the position of the $i$ body with respect \textbf{to the center of Jupiter}
\item $r_{ij}$ the distance between bodies $i$ and $j$
\item $R_J$ the radius of Jupiter
\item $J_n$ the dynamic polar oblateness of the nth order for Jupiter's gravity field
\item $U_{\bar{l}\hat{J}}$ potential generated by the oblateness of Jupiter on the satellite $l$
\item $\Phi_i$ is the latitude of the $i$ satellite with respect to Jupiter's equator.
\end{itemize}

@@

\begin{figure}
\includegraphics[width=8.8cm]{figures/Carme_ephemeris.eps}
\caption{Offsets in declination of the positions published by \protect\citetalias{GomesJunior2015} for Carme. The red "x" relative to the special-tailored ephemeris, the blue "+" relative to the jup300 JPL ephemeris and the green dot relative to \protect\cite{Emelyanov2005}. As expected, the ephemeris systematic errors pointed out by \protect\citetalias{GomesJunior2015} are reduced with the STE ephemeris. \label{Fig: JPL-STE}}
\end{figure}

In Table \ref{Tab:STE-error} it is presented the mean offsets and the respective standard deviation of the \citetalias{GomesJunior2015} positions relative to the same 3 ephemeris as above. We can see that the mean offsets as well as most of their standard deviations are greatly reduced with the STE ephemeris.

@@

For the specific case of Phoebe, the ninth satellite of Saturn, we have updated the ephemeris published in \cite{Desmars2013}. The new ephemeris (PH15) used the same dynamical model, including the perturbations of the Sun and the eight planets, the eight major satellites of Saturn and the $J_2$ parameter. The observations used to fit the model are identical to \cite{Desmars2013} (including 223 Cassini observations) with additional observations from \citetalias{GomesJunior2015}, \cite{Peng2015}, observations from Minor Planet Circulars between 2012 and 2014 (available on the Natural Satellite Data Center\footnote{\url{http://lnfm1.sai.msu.ru/neb/nss/bsapoouf.htm}}), and observations from Flagstaff \citep{NOFS} between 2012 and 2014. It represents a total number of 5886 observations from 1898 to 2014. In contrast, in \cite{Desmars2013} was used 3367 observations from 1898 to 2012. This represents an increase of almost 75\% in the number of observations, mainly with recent observations which is required for our purpose.

@@

We computed the precision of the PH15 ephemeris considering three sets of observations: all the positions available; only the positions of \citetalias{GomesJunior2015}; and all the positions without \citetalias{GomesJunior2015}. In Fig. \ref{Fig:err-Phoebe} we show the comparison between them for the time span 2016-2021 in angular separation\footnote{\label{ft-angsep}The precision in angular separation is computed  by $\sigma_s = \sqrt{\sigma^2_{\alpha} \cos^2 \delta + \sigma^2_{\delta}}$ where $\sigma_{\alpha}$ and $\sigma_{\delta}$ are the precision in right ascension and declination.} of Saturn. It is possible to see that even considering only the positions of \citetalias{GomesJunior2015}, the estimated error of the ephemeris is smaller than 12 mas. The computed precision does not take into account the precision in the position of Saturn.

\begin{figure}
\begin{centering}
\includegraphics[width=8.8cm]{figures/PH15_err_angsep.eps}  
\caption{Comparison of the precision in angular separation\textsuperscript{\ref{ft-angsep}} of Saturn for the PH15 ephemeris where three different sets of positions were used to compute the ephemeris: all observations; only positions of \citetalias{GomesJunior2015}; and all the positions without \citetalias{GomesJunior2015}.}
\label{Fig:err-Phoebe}
\end{centering}
\end{figure}

@@

A total of \noccs events were identified between January 2016 and December 2020. In Table \ref{Tab: satellite-occultation} we present the number of stellar occultations predicted by year for each satellite. It is possible to see an increase in the number of events found for Phoebe in 2018 and for the satellites of Jupiter in 2019-2020. This is because at that periods these satellites will cross the apparent galactic plane. We call attention that about 10\% of the events will involve stars brighter than magR=14 (and almost 25\% brighter than magR=15), which helps the attempt of amateur observers.

@@

Table \ref{Tab: sample-cds} shows a sample of the catalogue of predicted occultations and their parameters, which are necessary to produce occultation maps. Since these objects are very small, the duration of each event is a few seconds. All the occultation tables and maps will be publicly available at the CDS. In Fig. \ref{Fig: ocultacao} we show an example of an occultation map. This is an occultation by Elara that will happen in February 21, 2017. This event can be observed from Australia and it is one of the best opportunities for this object due to the slow velocity of the event and it involves a bright star (MagR*=12.4).

@@

\begin{figure}
\begin{centering}
\includegraphics[width = 8.8cm]{figures/Elara_2017-02-21T15:28:40.png}   
\caption{Occultation map for Elara. % regarding to the first event sampled in Table \ref{Tab: occ-list}.
The central red dot show the geocentric closest approach of the shadow. The small ones shows the center of the shadow separated by 60s. The lines show the path of the shadow over the Earth. The shadow moves from right to left.
\textbf{Labels:} Diam: Diameter of the object; Tmax: Maximum duration of the event for a central observation; C/A: the geocentric closest approach, in arcseconds; P/A: the satellite position angle with respect to the occulted star at C/A, in degrees; vel: velocity of event in km/s; Delta: Geocentric distance to the occulting object in AU; $R^*$: normalized magnitude to a common shadow velocity of 20 km s$^{-1}$; long: east longitude of subplanet point in degrees, positive towards east.}
\label{Fig: ocultacao}
\end{centering}
\end{figure}

@@

We performed new numerical integration for improving the orbits of some of the larger irregular satellites. Consequently, with more precise ephemeris, we predicted stellar occultations aiming to access fundamental parameters like size, shape, albedo, ultimately aiming to track the formation origin of these bodies.

For the irregular satellites of Jupiter (Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae and Sinope), we produced ephemeris using only the observations of \citetalias{GomesJunior2015}. The systematic errors found in the JPL ephemeris \citep{Jacobson2012} and \cite{Emelyanov2005} were improved. These new ephemerides are denominated Special-Tailored Ephemeris.

@@

As it was shown for Phoebe, when we use only the positions of \citetalias{GomesJunior2015}, the ephemeris presents a precision in the order of those where all the positions were used. Moreover, the case of Phoebe is particular because we have many observations in a large time spam (1898 to 2014) including observations from Cassini. For the Jovian satellites with less observations, the precision we have by using only \citetalias{GomesJunior2015} observations would be quite equivalent or even better than the precision of other ephemeris for the short time span explored in this work.

@@

\citetalias{GomesJunior2015} also observed Sycorax (satellite of Uranus) and Nereid (satellite of Neptune). Sycorax had a few observations on 9 nights in two years which did not cover one orbital period. For Nereid, the observations covered many orbital periods, but due to Nereid's large orbital eccentricity there are no observations near the pericenter.

>>>>

\usepackage[normalem]{ulem} %%% para mostrar o texto riscado

@@

\title[Orbits and predictions for occultations up to 2020]{New short-term orbits of irregular satellites in order to predict stellar occultations up to 2020}
%\title[Orbits and predictions for occultations up to 2020]{New orbits of irregular satellites based on thousands of new observations and predictions for stellar occultations up to 2020}
\author[A. R. Gomes-J\'unior, M. Assafin, L. Beauvalet et al.]{A. R. Gomes-J\'unior$^{1}$\thanks{E-mail: altair08@astro.ufrj.br},
M. Assafin$^{1,\dag}$,
L. Beauvalet$^{2,3}$,
J. Desmars$^{4}$,\newauthor
R. Vieira-Martins$^{1,2,5,\dag}$,
J. I. B. Camargo$^{2,5}$,
B. E. Morgado$^{1,2}$
F. Braga-Ribas$^{2,6}$
\\
$^{1}$Observat\'orio do Valongo/UFRJ, Ladeira Pedro Ant\^onio 43,
CEP 20.080-090 Rio de Janeiro - RJ, Brazil\\
$^{2}$Observat\'orio Nacional/MCTI, R. General Jos\'e Cristino 77, CEP 20921-400 Rio de Janeiro - RJ, Brazil\\
$^{3}$Observatoire de Paris/SYRTE, 77 Avenue Denfert Rochereau 75014 Paris, France\\
$^{4}$Institut de m\'ecanique c\'eleste et de calcul des \'eph\'em\'erides - Observatoire de Paris, UMR 8028 du CNRS, 77 Av. Denfert-Rochereau,\\ 75014 Paris, France\\
$^{5}$Laborat\'orio Interinstitucional de e-Astronomia - LIneA, Rua Gal. Jos\'e Cristino 77, Rio de Janeiro, RJ 20921-400, Brazil\\
$^{6}$Federal University of Technology - Paran\'a (UTFPR / DAFIS), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil
}
\begin{document}
\newcommand{\noccs}{5442 } % total number of occultations
\defcitealias{GomesJunior2015}{GJ15}

@@

We use the following notations: \begin{itemize}
\item \textbf{in one dynamical family consisting of $\mathcal{N}$ irregular satellites, $i$ will stand for the one whose equation of motion we are considering, $l$ will stand for another irregular satellite in gravitationnal interaction belonging to the same family}
\item $J$ Jupiter
\item $j$ another body of the Solar System, \textbf{among the 10 included (Sun+planets+Pluto+Moon)}
\item $M_j$ the mass of the $j$-th body, not an irregular satellite
\item $m_i$ the mass of the irregular satellite $i$
\item $\vec{r_i}$ the position of the $i$-th body with respect \textbf{to the center of Jupiter}
\item $r_{ij}$ the distance between bodies $i$ and $j$
\item $R_J$ the radius of Jupiter
\item $J_n$ the dynamic polar oblateness of the nth order for Jupiter's gravity field
\item $U_{\bar{l}\hat{J}}$ potential generated by the oblateness of Jupiter on the satellite $l$
\item $\Phi_i$ is the \textbf{latitude} of the $i$-th satellite with respect to Jupiter's equator.
\end{itemize}

@@

\begin{figure}
\includegraphics[width=8.8cm]{figures/Carme_ephemeris.eps}
\caption{Offsets in declination of the positions published by \protect\citetalias{GomesJunior2015} for Carme. The red "x" relative to the special-tailored ephemeris \textbf{(rms = 51)}, the blue "+" relative to the jup300 JPL ephemeris \textbf{(rms = 130)} and the green dot relative to \protect\cite{Emelyanov2005} \textbf{(rms = 92)}. As expected, the ephemeris systematic errors pointed out by \protect\citetalias{GomesJunior2015} are reduced with the STE ephemeris. \label{Fig: JPL-STE}}
\end{figure}

In Table \ref{Tab:STE-error} it is presented the mean offsets and the respective standard deviation of the \citetalias{GomesJunior2015} positions relative to the same 3 ephemeris as above. We can see that the mean offsets as well as most of their standard deviations are greatly reduced with the STE ephemeris. \textbf{It is obvious because the STE were based on the same observations with which they are compared. In contrast, two other ephemeris
were based on different sets of observations.}

@@

For the specific case of Phoebe, the ninth satellite of Saturn, we have updated the ephemeris published in \cite{Desmars2013}. The new ephemeris (PH15) used the same dynamical model, including the perturbations of the Sun and the eight planets, the eight major satellites of Saturn and the $J_2$ parameter. The observations used to fit the model are identical to \cite{Desmars2013} (including 223 Cassini observations) with additional observations from \citetalias{GomesJunior2015}, \cite{Peng2015}, observations from Minor Planet Circulars between 2012 and 2014 (available on the Natural Satellite Data Center \textbf{\citep{Arlot2009}}, and observations from Flagstaff \citep{NOFS} between 2012 and 2014. It represents a total number of 5886 observations from 1898 to 2014. In contrast, in \cite{Desmars2013} was used 3367 observations from 1898 to 2012. This represents an increase of almost 75\% in the number of observations, mainly with recent observations which is required for our purpose.

@@

We computed the precision of the PH15 ephemeris considering three sets of observations: all the positions available; only the positions of \citetalias{GomesJunior2015}; and all the positions without \citetalias{GomesJunior2015}. In Fig. \ref{Fig:err-Phoebe} we show the comparison between them for the time span 2016-2021 in on-sky Phoebe-Saturn angular separation\footnote{\label{ft-angsep}The precision in angular separation is computed  by $\sigma_s = \sqrt{\sigma^2_{\alpha} \cos^2 \delta + \sigma^2_{\delta}}$ where $\sigma_{\alpha}$ and $\sigma_{\delta}$ are the precision in right ascension and declination.}. It is possible to see that even considering only the positions of \citetalias{GomesJunior2015}, the estimated error of the ephemeris is smaller than 12 mas. The computed precision does not take into account the precision in the position of Saturn.

\begin{figure}
\begin{centering}
\includegraphics[width=8.8cm]{figures/PH15_err_angsep.eps}  
\caption{Comparison of the precision in on-sky Phoebe-Saturn angular separation\textsuperscript{\ref{ft-angsep}} for the PH15 ephemeris where three different sets of positions were used to compute the ephemeris: all observations; only positions of \citetalias{GomesJunior2015}; and all the positions without \citetalias{GomesJunior2015}.}
\label{Fig:err-Phoebe}
\end{centering}
\end{figure}

@@

A total of \noccs events were identified between January 2016 and December 2020. In Table \ref{Tab: satellite-occultation} we present the number of stellar occultations predicted by year for each satellite. It is possible to see an increase in the number of events found for Phoebe in 2018 and for the satellites of Jupiter in 2019-2020. This is because at that periods these satellites will cross the apparent galactic plane. We call attention that about 10\% of the events will involve stars brighter than magR=14  (and almost 25\% brighter than magR=15), \textbf{where magR stands for the UCAC4 magnitude in the R-band}, which helps the attempt of amateur observers.

@@

Table \ref{Tab: sample-cds} shows a sample of the catalogue of predicted occultations and their parameters, which are necessary to produce occultation maps. Since these objects are very small, the duration of each event is a few seconds. All the occultation tables and maps will be publicly available at the CDS \textbf{(Centre de Données astronomiques de Strasbourg)}. In Fig. \ref{Fig: ocultacao} we show an example of an occultation map. This is an occultation by Elara that will happen in February 21, 2017. This event can be observed from Australia and it is one of the best opportunities for this object due to the slow velocity of the event and it involves a bright star (MagR*=12.4).

@@

\begin{figure}
\begin{centering}
\includegraphics[width = 8.8cm]{figures/Elara_2017-02-21T15:28:40.png}   
\caption{Occultation map for Elara. % regarding to the first event sampled in Table \ref{Tab: occ-list}.
The central red dot shows the geocentric closest approach of the shadow. The small ones show the center of the shadow separated by 60s. The lines show the path of the shadow over the Earth. The shadow moves from right to left.
\textbf{Labels:} Diam: Diameter of the object; Tmax: Maximum duration of the event for\textbf{ a observation in the center of the shadow}; C/A: \textbf{apparent geocentric distance between the satellite and the star (a.k.a. the distance between the shadow and the center of the Earth) at the moment of the geocentric closest approach}, in arcseconds; P/A: the satellite position angle with respect to the occulted star at C/A, in degrees; vel: \textbf{relative} velocity of event in km/s; Delta: Geocentric distance to the occulting object in AU; $R^*$: normalized \textbf{UCAC4 magnitude in the R-band} to a common shadow velocity of 20 km s$^{-1}$; long: east longitude of subplanet point in degrees, positive towards east, \textbf{at the instant of the geocentric closest approach}.}
\label{Fig: ocultacao}
\end{centering}
\end{figure}

@@

We performed new numerical integration for improving the orbits of some of the larger irregular satellites. Consequently, with \textbf{our} ephemeris, we predicted stellar occultations aiming to access fundamental parameters like size, shape, albedo, ultimately aiming to track the formation origin of these bodies.

For the irregular satellites of Jupiter (Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae and Sinope), we produced ephemeris using only the observations of \citetalias{GomesJunior2015}. \textcolor{red}{The systematic errors found in the JPL ephemeris \citep{Jacobson2012} and \cite{Emelyanov2005} were improved.} These new ephemerides are denominated Special-Tailored Ephemeris.

@@

As it was shown for Phoebe, when we use only the positions of \citetalias{GomesJunior2015}, the ephemeris presents a precision in the order of those where all the positions were used. Moreover, the case of Phoebe is particular because we have many observations in a large time spam (1898 to 2014) including observations from Cassini. For the Jovian satellites with less observations, the precision we have by using only \citetalias{GomesJunior2015} observations \textbf{may} be quite equivalent or even better than the precision of other ephemeris for the short time span explored in this work.

@@

\citetalias{GomesJunior2015} also observed Sycorax (satellite of Uranus) and Nereid (satellite of Neptune). \textbf{There were few observations of Sycorax distributed in 9 nights over two years} which did not cover one orbital period. For Nereid, the observations covered many orbital periods, but due to Nereid's large orbital eccentricity there are no observations near the pericenter.
